---
title: "Challenge 7: Incorporating Multiple Inputs"
subtitle: "Functions + Fish"
author: "Ryan Chan"
format: 
  html:  
    embed-resources: true
    code-tools: true
    toc: true
    theme: lux
code-fold: true
editor: source
execute: 
  echo: true
  warning: false
  error: true
---

```{r}
#| label: setup

library(here)
library(tidyverse)

```

```{r}
#| label: load-data

fish <- read_csv(here::here("data",
                            "BlackfootFish.csv"),
                 show_col_types = FALSE)
```
## Condition Index

A frequently used measurement for fish health is a condition index [(Wikipedia article)](https://en.wikipedia.org/wiki/Condition_index_in_fish). The following
simple equation can be used to calculate the approximate condition index of a
fish:

$$\text{condition index} = \frac{weight}{length^3} \times 100$$

**1. There are specific units required for the calculation of a condition index -- length must be in centimeters and weight must be in grams. The weight data for the Blackfoot River fish were collected in grams; the length data were collected in millimeters. Transform the length data to the correct units.**

```{r}
#| label: transform-data-to-have-correct-units

fish <- fish |>
  mutate(length = length / 10) # divide by 10 to convert to cm

```

**2. Collecting data of this sort can be very messy! Write a function that will replace unlikely `length` and `weight` measurements with `NA`. Your function should accept at least three inputs:**

+ **a vector of measurements,**
+ **the minimum reasonable value,**
+ **the maximum reasonable value.**

**If a value falls outside these bounds, you should replace it with an `NA`.**

::: callout-tip
If you are struggling with the structure of your function, I would suggest reading the [Mutating Function](https://r4ds.hadley.nz/functions.html#mutate-functions) from R4DS.
:::

```{r}
#| label: function-to-change-impossible-measurements-to-NA

filter_measurements <- function(vec, lower, upper){
  return(ifelse(vec >= lower & vec <= upper,
                vec,
                NA))
}

```

**3. After consulting the [Montana Record Table](https://fwp.mt.gov/fish/anglingData/records) for the four species of trout included in these data, I have conjectured that it is unlikely to have measurements for fish below 5 cm and above 80 cm in length or below 10 g and above 4,000 g in weight. Use your function to modify the `length` and `weight` columns of the `fish` dataset based on my cutoffs.**

```{r}
#| label: modify-impossible-values-of-length-weight

fish <- fish |>
  mutate(length = filter_measurements(length, lower = 5, upper = 80),
         weight = filter_measurements(weight, lower = 10, upper = 4000))

```

**4. Write a function to calculate the condition index of a fish, given inputs of weight and length.**

::: callout-warning
Consider whether you will write a function that works with vectors (accepting
vectors as inputs) or a function that works with data frames (accepting a 
dataset and variable names as inputs)!
:::

```{r}
#| label: condition-function

condition_index <- function(length, weight){
  return(100 * (weight / (length^3)))
}

```

**5. Make a thoughtful visualization of how fish conditions have varied over the duration of this study.**

::: callout-warning
### To be thoughtful...

... you need to use at least a third variable!
:::

```{r}
#| label: condition-indices-over-time

fish |>
  # calculate condition indices for fish
  mutate(condition_index = condition_index(length,
                                           weight)) |>
  filter(!is.na(condition_index)) |>
  # plot condition index over time by species
  ggplot(aes(x = year, 
             y = condition_index, 
             color = species)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(method = "loess", 
              se = FALSE, 
              size = 0.8) +
  labs(title = "Blackfoot River Trout Condition Index Over Time by Species",
       x = "Year",
       y = "",
       subtitle = "Condition Index",
       color = "Species") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~ species, 
             scales = "free_y")  # facet by species
```
Across the study period from 1989 to 2006, condition indices for all four trout species in the Blackfoot River have remained relatively stable. For brown trout, the condition index shows moderate variability with a higher condition index value, with most values concentrated between 0 and 10. Bull trout look to have a smaller range in condition index values, primarily between 0.5 and 2. This demonstrates an even more consistent condition over time with minimal fluctuations. Rainbow trout exhibit the highest variability of condition over time, with condition indices reaching values above 30. Despite the high variability, there doesn't seem to be an apparent trend over time. Finally, the West Slope Cutthroat Trout also show some variability, although less prominent than rainbow trout, and their trend implies stable overall conditions. 
